---
title: "Buffering of transcription rate shifts"
output:
  html_notebook:
    toc: true
    number_sections: true
    theme: readable
author: Marat Mufteev
date: "`r format(Sys.time(), 'Generated on %d %B, %Y')`"
---

Current approach in the estimate of transcription rate (TR) shifts buffering suffers from the interdependence of TR and half-life (HL) measurement. Indeed, the denominator in a ratio method is TR itself.

Denote measured TR and steady-state (SS) log2 fold-change as logT and logS. Similarly denote real TR and steady-state (SS) log2 fold-change as logT_r and logS_r. Finally, let N_t and N_s be noise components for the TR and SS respectivelty. Then:

logT = logT_r + N_t
logS = logS_r + N_s

Estimated HL log2 fold-change logHL is:
logHL = logS - logT = (logS_r - logT_r) + (N_s - N_t) = logHL_r + (N_s - N_t)

Then, in a heatmap we look at a comparison between logT and logHL. There are two main regimes.

A) Noise is bigger than a real signal (N_s - N_t >> logHL_r & N_t >> logT_r). Then we basically compare:
N_t vs (N_s - N_t)
In this case, a noise induced negative relation between TR and HL changes will be observed.

B) Signal is bigger than noise (logT_r and logHL_r dominates).
Then, we observe true signal logT_r vs logHL_r

A clean description of buffering would combine "pulse and chase" approach of HL measurement (when decay of labeled mRNAs is observed) with our current approach of TR measerement.

Upload libraries
```{r}
library(dplyr)
library(ggplot2)
```

Set the path to data directory
```{r}
root = "./data/"
```

Upload transcription rate, steady state and half-life data
```{r}
TR_raw = read.csv(paste(root, "TR_log2FC_deseq2.csv.gz", sep = ""), header = T)
SS_raw = read.csv(paste(root, "SS_log2FC_deseq2.csv.gz", sep = ""), header = T)

differentiation_step = "NPC_iPS"

# We also tried to estimate SS/TR fold change with a DESeq2
# from interaction term celltype:assay in a design:
# ~ celltype + replicate + assay + celltype:assay
# Here assay is a factor with values c(SS, TR)
SS_TR_raw = read.csv(paste(root, "SS_TR_deseq2.csv.gz", sep = ""), header = T)

SS_TR_tbl = rbind(SS_TR_raw %>% 
                    mutate(log2FC_SS_TR = log2_SS_TR_NPC_WT - log2_SS_TR_iPS_WT,
                           transition = "NPC_iPS"),
                  SS_TR_raw %>% 
                    mutate(log2FC_SS_TR = log2_SS_TR_Neu_WT - log2_SS_TR_NPC_WT,
                           transition = "Neu_NPC")) %>%
            filter(transition == differentiation_step) %>%
            select(gene, log2FC_SS_TR)


normed_avg_tbl = read.csv(paste(root, "normed_avg_counts.csv.gz", sep = ""), header = T) %>%
                 select(gene, contains("SS"))
```

Define thresholds for a sufficient log2FC
```{r}
TR_tbl = TR_raw 
colnames(TR_tbl) = gsub(differentiation_step, "TR", colnames(TR_tbl))
TR_tbl = TR_tbl %>%
         mutate(padj_TR = ifelse(is.na(padj_TR), 1, padj_TR),
                TR_status = "Not",
                TR_status = ifelse(log2FC_TR > 0 & padj_TR < 0.1,
                                       "Up", TR_status),
                TR_status = ifelse(log2FC_TR < 0 & padj_TR < 0.1,
                                       "Down", TR_status)) %>%
         select(gene, log2FC_TR, TR_status)

SS_tbl = SS_raw 
colnames(SS_tbl) = gsub(differentiation_step, "SS", colnames(SS_tbl))
SS_tbl = SS_tbl %>%
         mutate(padj_SS = ifelse(is.na(padj_SS), 1, padj_SS),
                SS_status = "Not",
                SS_status = ifelse(log2FC_SS > 0 & padj_SS < 0.1,
                                       "Up", SS_status),
                SS_status = ifelse(log2FC_SS < 0 & padj_SS < 0.1,
                                       "Down", SS_status)) %>%
         select(gene, log2FC_SS, SS_status)

RATE_tbl = full_join(TR_tbl, SS_tbl) %>%
           select(gene, log2FC_TR, log2FC_SS, TR_status, SS_status)
```

Tidy log2FC data for a heatmap
```{r}
RATE_clean = RATE_tbl %>%
              mutate(TR_na = is.na(log2FC_TR),
                     SS_na = is.na(log2FC_SS),
                     log2FC_SS = ifelse(SS_na & !TR_na,
                                        0, log2FC_SS),
                     SS_status = ifelse(SS_na & !TR_na,
                                            "Not", SS_status),
                     log2FC_TR = ifelse(TR_na & !SS_na,
                                        0, log2FC_TR),
                     TR_status = ifelse(TR_na & !SS_na,
                                            "Not", TR_status)) %>%
              na.omit() %>%
              filter(TR_status != "Not" | 
                     SS_status != "Not") %>%
              # Reduce fold-change value by 1/2 for visual purpose
              mutate(log2FC_TR = ifelse(TR_status == "Not",
                                        0.5*log2FC_TR,
                                        log2FC_TR),
                     log2FC_SS = ifelse(SS_status == "Not",
                                        0.5*log2FC_SS,
                                        log2FC_SS)) %>%
              select(-TR_na, -SS_na)

# Reshape the data
RATE_melt = reshape2::melt(RATE_clean,
                           id.vars = c("gene", "TR_status", "SS_status")) %>%
            mutate(variable = gsub("log2FC_", "", variable),
                   variable = factor(variable,
                                     levels=c("TR", "SS")))
```

Define heatmap groups
```{r}
heatmap_data = RATE_melt %>%
    mutate(variable = gsub("log2FC_", "", variable),
           variable = factor(variable,
                             levels=c("TR", "SS"))) %>%
    group_by(gene) %>%
    mutate(SS_TR_status = ifelse(abs(value[variable == "SS"] - value[variable == "TR"])/abs(value[variable == "TR"]) > 0.25, 
                                 "Yes", "Not"),
           SS_TR_status = ifelse(SS_TR_status == "Not" & 
                                   ((TR_status == "Up" & SS_status %in% c("Not", "Down")) | 
                                      (TR_status == "Down" & SS_status %in% c("Not", "Up"))), 
                                 "Yes", SS_TR_status),
           SS_TR_status = ifelse(value[variable == "SS"] - value[variable == "TR"] > 0 & SS_TR_status == "Yes", 
                                 "Yes_SS", SS_TR_status),
           SS_TR_status = ifelse(value[variable == "TR"] - value[variable == "SS"] > 0 & SS_TR_status == "Yes", 
                                 "Yes_TR", SS_TR_status)) %>%
    ungroup() %>%
    mutate(reg_group = ifelse(SS_TR_status == "Not", "Follow", NA),
           reg_group = ifelse((TR_status == "Up" & SS_TR_status == "Yes_SS") | 
                                (TR_status == "Down" & SS_TR_status == "Yes_TR") | 
                                (TR_status == "Not" & (SS_TR_status %in% c("Yes_SS", "Yes_TR"))), 
                              "Boost", reg_group),
           reg_group = ifelse((TR_status == "Up" & SS_TR_status == "Yes_TR") | 
                              (TR_status == "Down" & SS_TR_status == "Yes_SS"), 
                              "Buffer", reg_group),
           reg_group = ifelse(reg_group=="Boost" & 
                              ((TR_status == "Down" & SS_status == "Not") |
                               (TR_status == "Up" & SS_status == "Not")), 
                              "Buffer", reg_group),
           reg_group = ifelse(reg_group == "Boost" & TR_status == "Not", 
                              "Boost_HL", reg_group),
           SS_status = ifelse(reg_group == "Buffer" & TR_status == "Down" & SS_status == "Down", 
                              "Not", SS_status),
           SS_status = ifelse(reg_group == "Buffer" & TR_status == "Up" & SS_status == "Up", 
                              "Not", SS_status),
           SS_status = ifelse(reg_group=="Buffer" & TR_status == "Down" & SS_status == "Up", 
                              "Not", SS_status),
           SS_status = ifelse(reg_group=="Buffer" & TR_status == "Up" & SS_status == "Down", 
                              "Not", SS_status),
           reg_group = factor(reg_group,
                              levels=c("Follow", "Boost", "Boost_HL", "Buffer"))) %>%
    mutate(value = (e1071::sigmoid(0.5*value) - 0.5))
```

Sort genes
```{r}
genes = (heatmap_data %>%
             filter(variable == "TR") %>%
             mutate(TR_status = factor(TR_status,
                                       levels = c("Down", "Not", "Up"))) %>%
             arrange(reg_group, TR_status, SS_status, value) %>%
             mutate(gene = factor(gene, levels=unique(gene))))$gene
heatmap_df = heatmap_data %>%
             mutate(gene = factor(gene, levels=genes),
                    TR_status = factor(TR_status,
                                       levels = c("Down", "Not", "Up")))
```

Visualize buffering heatmap
```{r}
borders = (heatmap_df %>%
               filter(variable == "TR") %>%
               group_by(reg_group, TR_status, SS_status) %>%
               summarise(n=n()))$n
  
output = paste(root, "heatmap_buffering.pdf", sep = "")
ggplot(data = heatmap_df,
         aes(x = variable,
             y = gene,
             fill = value)) +
    geom_tile() +
    geom_hline(yintercept = cumsum(borders)) +
    theme(axis.title.y=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank()) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    scico::scale_fill_scico(palette = "cork",
                            breaks = c(-0.4241418, -0.2772999, 0, 0.2772999, 0.4241418),
                            labels = c(-5, -2.5, 0, 2.5, 5),
                            name = latex2exp::TeX("$log_{2}$(FC)")) +
    theme(plot.background = element_rect(fill = "white",
                                         colour = "white",
                                         size = 1))
    ggsave(output, width = 1.5, height = 3)
```


Prepare data for the barplot of TR, SS and SS_TR fold-changes
```{r}
RATE_collector=list()
for (transition in c("NPC_iPS", "Neu_NPC")) {
  log2FC_var=paste("log2FC_", transition, sep = "")
  lfcSE_var=paste("lfcSE_", transition, sep = "")
  padj_var=paste("padj_", transition, sep = "")
  
  for (assay in c("TR", "SS")) {
    log2FC_out = paste("log2FC_", assay, sep = "")
    lfcSE_out = paste("lfcSE_", assay, sep = "")
    padj_out = paste("padj_", assay, sep = "")
    
    input_path = paste(root, assay, "_log2FC_deseq2.csv.gz", sep = "")
    output_tbl = read.csv(input_path, header = T) %>%
                 mutate(!!log2FC_out := .[,log2FC_var],
                        !!lfcSE_out := .[,lfcSE_var],
                        !!padj_out := .[,padj_var],
                        transition) %>%
                 select(gene, log2FC_out, lfcSE_out, padj_out, transition)
    RATE_collector[[paste(assay, transition, sep = "_")]] = output_tbl
  }
  
  input_path = paste(root, "SS_TR_deseq2.csv.gz", sep = "")
  cells = strsplit(transition, "_")[[1]]
  log2FC_varB = paste("log2_SS_TR_", cells[1], "_WT", sep="")
  log2FC_varA = paste("log2_SS_TR_", cells[2], "_WT", sep="")
  lfcSE_varB = paste("log2_SS_TR_", cells[1], "_WT", sep="")
  lfcSE_varA = paste("log2_SS_TR_", cells[2], "_WT", sep="") 
  RATE_SS_TR = read.csv(input_path, header = T) %>%
               mutate(log2FC_SS_TR = .[,log2FC_varB] - .[,log2FC_varA],
                      lfcSE_SS_TR = sqrt(.[,lfcSE_varB]**2 + .[,lfcSE_varA]**2)/sqrt(4),
                      Z_SS_TR = abs(log2FC_SS_TR)/sqrt(lfcSE_SS_TR),
                      transition) %>%
               select(gene, log2FC_SS_TR, lfcSE_SS_TR, Z_SS_TR)
  RATE_collector[[paste("SS_TR", transition, sep = "_")]] = RATE_SS_TR
}

transition_step = "NPC_iPS"
RATE_FULL = full_join(RATE_collector[[paste("SS_", transition_step, sep="")]], 
                      RATE_collector[[paste("TR_", transition_step, sep="")]], 
                      by = c("gene", "transition")) %>%
            left_join(RATE_collector[[paste("SS_TR_", transition_step, sep="")]])
```

Select genes
```{r}
if (transition_step == "NPC_iPS") {
  genes = c("MYT1", "MSI2", "NDUFAF7",
          "SETD2", "FEZ2", "EIF4E3",
          "NRXN2", "HNRNPF", "VEGFA",
          "SYP", "CTNND2", "DDX25")
} else {
  genes = c("ELAVL4", "SMAD3", "SYN1",
          "DSN1", "EXOSC9", "DLG4",
          "PEX11G", "FBXL2", "TDRD3", 
          "ZSWIM5", "FOXP1", "DKK3")
}
```

Format data for barplot
```{r}
df_A = reshape2::melt(RATE_FULL %>%
                        filter(gene %in% genes) %>%
                        select(gene, contains("log2FC")) %>%
                        select(-contains("vs"))) %>%
       mutate(assay = gsub("log2FC_", "", variable),
              variable = gsub("_SS|_TR", "", variable),
              MEAN = value) %>%
       select(gene, assay, MEAN)
df_B = reshape2::melt(RATE_FULL %>%
                        filter(gene %in% genes) %>%
                        select(gene, contains("lfc")) %>%
                        select(-contains("vs"))) %>%
       mutate(assay = gsub("lfcSE_", "", variable),
              variable = gsub("_SS|_TR", "", variable),
              SE = value) %>%
       select(gene, assay, SE)

reg_tbl = data.frame(gene = genes,
                     reg_type = factor(c(rep("Transcriptin only", 3), 
                                         rep("RNA stability only", 3),
                                         rep("Buffering", 3), 
                                         rep("Boosting", 3)),
                                       levels=c("Transcriptin only", 
                                                "RNA stability only", 
                                                "Buffering", "Boosting")))
df = full_join(df_A, df_B) %>%
     left_join(reg_tbl) %>%
     #filter(assay != "HL") %>%
     #bind_rows(df_C) %>%
     mutate(assay = factor(assay, levels = c("TR", "SS", "SS_TR")),
            gene = factor(gene, levels=genes))
```

Show barplot
```{r}
output = paste(root, "barplot_buffering.pdf", sep = "")
position = position_dodge()
ggplot(data = df,
       aes(x = gene,
           y = MEAN,
           fill = assay)) +
  geom_bar(stat="identity", position=position, width = 0.7) +
  geom_errorbar(aes(ymin = MEAN - SE,
                    ymax = MEAN + SE), width = 0.2, 
                position = position_dodge(width=0.7)) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = c(1.5, 2.5), linetype = 5, alpha = 0.8) +
  ggthemes::theme_base(base_size = 10) +
  scale_fill_manual(values=c("#23C897", "#8761B0", "#FFA568")) +
  facet_wrap(~reg_type, nrow = 1, scales = "free_x") +
  coord_cartesian(ylim = c(-3.4, 3.4)) +
  ylab(latex2exp::TeX("$log_{2}$ fold-change")) +
  theme(plot.background = element_rect(fill = "white", colour = "white", size = 1),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
        strip.text.x = element_text(angle = 0))  
  ggsave(output,
         width = 7, height = 2.48)
```



