---
title: "Global average half-life per cell type"
output:
  html_notebook:
    toc: true
    number_sections: true
    theme: readable
author: Marat Mufteev
date: "`r format(Sys.time(), 'Generated on %d %B, %Y')`"
---

Upload libraries
```{r}
library(dplyr)
library(ggplot2)
```

Set the path to data directory
```{r}
root = "./data/"
```

Upload:
A) half-life (HL) data from a saturation curve fit
B) yield quantifications from cellular fractions
```{r}
HL_tbl = read.csv(paste(root, "HL_saturation_curve.csv.gz", sep = ""),
                  header = T) %>%
         mutate(celltype = gsub("_WT", "", celltype),
                celltype = factor(celltype,
                                      levels=c("iPS", "NPC", "Neu")))

yield_data = read.csv(paste(root, "RNA_extraction_cellular_fractions.csv", sep = "")) %>%
             tidyr::separate(Sample, c("celltype", "fraction"), " ") %>%
             group_by(celltype, fraction) %>%
             mutate(rep = 1:n()) %>%
             ungroup() %>%
             mutate(fraction = gsub("total", "Total", fraction),
                    fraction = factor(fraction,
                                      levels=c("Nuc", "Cyto", "Total")),
                    celltype = factor(celltype,
                                      levels=c("iPS", "NPC", "Neu")))
```

To measure average half-life (HL) we attempted multiple methods.
Genes included for the calculation:
A) all genes with measured HL
B) unreliable genes are removed (degree of stringency could be varied)
C) HL values for unreliable genes are interpolated from a ratio method
We noticed that the conclusion on a global HL shift didn't depend on the method.
```{r}
# Method B 
# Thresholds are somewhat arbitrary
# However, the higher average HL in RTT neurons is not too sensitive to the thresholds in a reasonable range
HL_clean = HL_tbl %>%
           group_by(celltype) %>%
           filter(TR > quantile(TR, 0.1, na.rm = T) | 
                  SS > quantile(SS, 0.1, na.rm = T),
                  CI50_HL < 0.75) %>%
           ungroup() %>%
           select(gene, celltype, HL)
```

Visualize a global shift in HL
```{r}
output = paste(root, "global_mRNA_HL.pdf", sep = "")
ggplot(data = HL_clean,
       aes(x = log(HL, 10),
           color = celltype)) +
  stat_density(geom = "line", position = "identity", size = 1) +
  scale_color_manual(values = c("#00FF95", "#FF0000", "#0092FF", "#FFAA00")) +
  ggthemes::theme_base(base_size = 25) +
  scale_x_continuous(breaks = c(0, 0.4771213, log(9, 10)), labels = c(1, 3, 9), 
                     limits = c(log(0.5, 10), log(24, 10))) +
  facet_wrap(~celltype, ncol = 1) +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        plot.background = element_rect(fill = "white",colour = "white",size = 1))
  ggsave(output, width = 4, height = 5)
```

Visualize shifts in cellular fractions RNA yield
```{r}
yield_stats = yield_data %>%
              group_by(celltype, fraction) %>%
              summarise(yield_avg = mean(Yield_RNA_cell * 1e6),
                        yield_se = sd(Yield_RNA_cell * 1e6)/sqrt(n()))

output = paste(root, "RNA_yield.pdf", sep = "")
ggplot(data = yield_stats,
       aes(x = celltype,
           y = yield_avg,
           group = fraction)) +
  geom_line(aes(color = fraction)) +
  geom_errorbar(aes(ymin = yield_avg - yield_se,
                    ymax = yield_avg + yield_se),
                width = 0.2) +
  geom_point(aes(fill = fraction), size = 2.5, pch = 21, color = "black") +
  ggthemes::theme_base() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        plot.background = element_rect(fill = "white", colour = "white", size = 1)) +
  ylab("") +
  scale_color_manual(values=c("#CC6666", "#9999CC", "#66CC99")) +
  scale_fill_manual(values=c("#CC6666", "#9999CC", "#66CC99")) 
  ggsave(output, width = 2.5, height = 2.5)
```




